\begin{acknowledgment}

I would like to thank my advisor, Dr.~Barry Williams,
for taking me under his mentorship.
%
Barry has been an excellent source of ideas
that have helped me frame and focus my dissertation.
%
He has been very supportive throughout my time here
and open to my exploration of new topics,
some of which turned out to be chapters in my dissertation.
%
Through the professional connections he has helped me make,
Barry has helped me further my scientific career.
%
Finally, Barry has been a friend as well as a colleague,
and through him I have learned much about life.



I would like to thank my dissertation committee,
Drs.~Richard Lenski, Charles Ofria, and Doug Schemske.
%
They have been incredibly helpful
in developing my ideas and focusing my research.
%
Listening to them discuss science in my committee meetings was always inspiring.
%
I would also like to thank Dr.~Luke Harmon,
who collaborated with me on one of my chapters.
%
Luke has been very positive and cared for my well-being
because he usually started our conversations with ``Are you having fun?''
%
I would like to thank my other collaborators,
who, by sharing their knowledge and wisdom,
have made me a better scientist:
Dr.~Jenny Boughman, Dr.~Alex Shingleton, Dr.~Jeff Barrick,
Dr.~Shampa Ghosh, Dr.~Lili Lettieri, Dr.~Jason Keagy,
Dr.~Bj\o rn \O stman, and Alycia Lackey.



I would like to thank the many excellent people of the Williams lab---%
past and present.
%
Vicki Rademacher, Kristine Cox, Krista Reitenga, and Siva Nallasivam
did a lot of the lab work that some of my research explorations relied upon.
%
Alex Prediger, Bahaar Chawla, Molly Roseland, and other undergraduates
helped me run experiments or kept the lab clean and well-supplied.
%
I thank my fellow graduate lab members, Krista Reitenga, Katy Califf,
Chris Hamm, Cory Kohn, and Michael Grispo.



I would like to thank the Devolab for their great feedback
during lab presentations that have greatly improved my research.
%
Several people (including friends named later) offered great comments
on parts of my work, especially Dr.~Ian Dworkin, Dr.~Aleeza Gerstein,
Dr.~Daniel Matute, Emily Weigel, and Dr.~Sonal Singhal.
%
I would also like to thank Dr.~Danita Brandt, Dr.~Michael Velbel,
Dr.~Michael Gottfried, Dr.~Robert Anstey, and Summer Ostrowski
for enriching my mind with great scientific discussions.



I would like to thank the many people who worked in the background
and made my life a little easier: Brian Baer, Pat Bills, Dirk Colbry, Rob Lyon,
Lisa Craft, Julia Ahmed, Pat Resler, and Connie James.
%
Dr.~Clifford Weil taught me \LaTeX\ and provided me with the template that
helped create this beautiful-looking dissertation.
%
My work would have been impossible without the many agencies and organizations
that funded me: Michigan State University, the Department of Zoology,
the Quantitative Biology Initiative, the Graduate School,
the Ecology, Evolutionary Biology and Behavior (EEBB) Program,
and the BEACON Center for the Study of Evolution in Action.



Throughout my six years here, I have made some amazing friends,
which alone has made this experience worthwhile.
%
I especially thank Eli Swanson, Austin Dreyer, and Nick Testa.
%
I would also like to offer warm thanks to Greg Stricker, Kelly Barnes,
Chelsea Van Eck, Darby Monks, Aaron Florn, Ben Dantzer, Adrian Dantzer,
Emily Dittmar, Katy Califf, Heather Stadden, Gaurav Moghe, Lorna Watt,
Mridul Thomas, Kane Keller, and Lori Torgerson-White.
%
Back home, I would like to thank my best friends,
Enrique Pereira and Duvan Cope,
and my high school physics teacher ``Mr.\ Moore,''
who was an inspiration to me and to whom I dedicate this dissertation.



Finally, I would like to thank my family.
%
My parents, Jesus and Crystel,
my sister Cristel, and my brothers Pablo and Marcelo
have given me love, support, and encouragement throughout my life.
%
I would also like to thank my grandparents and extended family,
especially my grandmother Zarela, who will soon turn 93,
for sharing her love and wisdom with me.
%
Finally, I would like to thank my little companions,
both past and present, who have done nothing but to simply be cute,
especially Baxter, Johnny, and Angel.

\end{acknowledgment}
