\begin{doublespace}

\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}

% Possible order:
% 1. sgv
%    - most adaptation happened through sgv
% 2. rate of compensation
%    - studied factors affecting compensation vs reversion
% 3. compensatory speciation
%    - can compensation lead to incompatibilities?
% 4. ecol vs MO
%    - ecol > MO; MO sensitive to gene flow
% 5. snowball
%    - quadratic increase in DMIs... maybe linear for hybrid incomp.

% Rough ideas for Introduction:
% - General: I think I want to emphasize that I'm using Avida
%   to answer question in adaptation and speciation that are difficult
%   to track with biological organism, and models are too simplistic.
% - First paragraph: Talk about genetics of adaptation
%   and mention that tracking individual mutations is important
% - Second paragraph: Talk about genetics of speciation
%   (emphisize on postzygotic) and that it is important
%   to understand those mutations that cause incompatibilities
% - Say that first I'm going to briefly overview how Avida works,
%   then say that I'm going to summarize my chapters

\section*{Background}
\addcontentsline{toc}{section}{Background}

Adaptation is the process in which beneficial traits
are acquired by organisms in a population through time.
%
This process is driven by natural selection.
%
Adaptive traits are ultimately encoded in the genetic material of organisms,
and therefore adaptation is often thought of in terms of adaptive alleles.
%
One of the goals of evolutionary research
is to understand the genetic basis of adaptation.
%
For example, do populations introduced to a new environment
adapt from new mutations or from ``standing genetic variation''
(i.e., allelic variation present in the population) \citep{orr05}?
%
Questions like this require detailed analyses of genetic data,
such as tracking individual alleles through time \citep{bar08}.



Speciation, the process by which new species form,
is often a by-product of adaptation \citep{coy04,sch09,sob10}.
%
Populations that evolve independently are likely to adapt in different ways
and therefore diverge genetically.
%
Genetic differences are likely to create ecogeographic, morphological,
physiological, or genetic isolation between populations \citep{coy04,sch10},
collectively known as ``reproductive isolation.''
%
For example, populations may not recognize each other as potential mates
or may produce sterile or inviable hybrids.
%
Many speciation researchers are interested
in the genetic mechanisms that lead to reproductive isolation.
%
Like in studies of the genetic basis of adaptation,
an understanding of the genetic basis of speciation
requires detailed analyses of genetic data.


% Written with Doug in mind, make an argument for Avida;
% discussion both of the limits of theory and smiualtion
% as well as the properties of this AL system that lends iteslf
% to biology -- cool to have a table to compare methods
%
% What Avida provides that other systems do not provide
% (think about how others describe the benefits of their own system).

%Detailed analysis of genetic data is difficult and time consuming.
%

Richard Feynman once said, ``If our small minds, for some convenience,
divide this ... universe, into parts---%
physics, biology, geology, astronomy, psychology, and so on---%
remember that nature does not know it.''
%
`Biology' is nothing more than a convienient term used to categorize
certain kinds of complex natural patterns and processes.
%
Many of these processes, however, are not phenomena
that strictly occur in living systems.
%
For example, the process of diffusion in the electron transport chain,
which generates energy for cells in the form of ATP molecules,
is used by biological systems but is not exclusively a biological process.
%
Diffusion is a natural phenomenon that occurs whenever
particles exhibit Brownian, or random, motion,
and was, in fact, first studied by physical scientists.
%
In the same way, evolution via natural selection occurs in biological systems,
but it is not exclusively a biological process \citep{pen07}.
%
Cultural phenomena, such as languages and ideas,
and artificial genetic systems, which I discuss later,
can also evolve via selection.
%
Natural selection is a universal process that occurs whenever
three conditions are met:
(1)~inheritance, (2)~variation, and (3)~differential survival \citep{ada06}.



Therefore, the process of evolution can be studied in isolation.
%
Mathematical models and computer simulations of evolutionary processes
show that it can be studied outside of the biological realm.
%
Using such models we have learned a great deal about evolution,
creating predictions that can then be tested in biological organisms.
%
However, models are limited in that they cannot capture
the complexity and open-endedness of evolution in action \citep{yed01}.
%
For example, models often deal with only a range of parameters values
and their possible outcomes is often limited in scope,
so that new patterns or new behavior is not possible \citep{yed01}.
%
Biological models, such as the use of \emph{E. coli} and \emph{S. cerevisiae},
have provided great insights about the working of evolution,
but even these systems can be intractable.
%
A middle-ground between these extremes---mathematical models
and biological ones---is the use of artificial life systems \citep{yed01}.
%
These are systems in which the conditions for evolution are present,
and therefore constitute an instance of evolution \citep{pen07}.

% First time anyone has studied specitaion (R.I.) in Avida

% Models & simulations
% - "digital life can help to refine mathematical theories and aid in
%   developing and quickly testing new hypotheses about ecological and
%   evolutionary processes. (Wilke 2002)
% - "‘I never intended [Tierra] as a model, but that’s the way a lot of people
%   saw it because they weren’t really prepared for this new idea, this
%   different perspective of another instance of life’, he says." [Ray] (O'Neil)

% General statements
%
% - "there could be more to life than we have so far assumed. Alifers are
%   challenging our common conceptions of life by posing questions whose
%   answers would be crucial to a comprehensive understanding of the nature of
%   life, yet ones that biologists themselves typically do not ask." (Swan)
% - ME: Use biology when you want to understand something fundamentally
%   biological, but use both when you want to understand general processes
% - "insights provided by s-Alife research have the potential to fundamentally
%   alter our understanding of the nature of organic life and thus deserve the
%   attention of both philosophers and natural scientists. (Swan)
% - a-life as a "thought experiment" (defined as "allowing us to entertain what
%   might be, or what could have been, given different parameters") (Swan)

% Benefits of Avida
%
% - Complete knowledge of individual organisms, their genotype, fitness,
%   ancestry; and can track individual alleles
% - digital "experiments will allow more detailed analyses of complex phenomena
%   and, in turn, might stimulate increasingly sophisticated evolution
%   experiments with real organ- isms." (Elena/Lenski 2003)
% - "‘It’s not just the speed, by any means’, says Lenski. ‘It’s also the power
%   to manipulate almost any variable one can imagine, to measure variables
%   with absolute precision, to store information that then allows one to trace
%   back a complex chain of events, and to take evolved organisms and subject
%   them to new sorts of analyses that one might not even have anticipated when
%   first collecting the data’. (O'Neil)

% Similarity w/ biology
% - "... `digital organisms' are not nealy as sophisticated as simple bacteria,
%   or even real viruses." (Twice as natural.)
% - "The physiology of these programs is different in many respects from that
%   of organic creatures, but their population biology has many features in
%   common, including a very extensive, if not unbounded, range of variation."
%   (Yedid, 2001).
% - "Although the physiology of Tierran organisms is bizarre, their
%   evolutionary dynamics seem likely to resemble those of microbes in most
%   essential respects. (Yedid, 2001)
% - "digital organisms provide a unique opportunity with which to study
%   evolutionary biology in a form of life that shares no ancestry with
%   carbon-based life forms, and hence to distinguish general principles of
%   evolution from historical accidents" (Wilke, 2002)
% - "‘But even if the digital and biological realms sometimes come into
%   scientific conflict, it would only lead one to ask why and then probe the
%   relevant factors more deeply’. (O'Neil)

% Limitations of biological methods
%
% - "current technology does not allow us to characterize the genotypes of
%   thousands of cells per generation, still less to determine with certainty
%   their ancestor-descendant relationships, except in the simplest systems,
%   such as phages and ribozymes" (Yedid, 2001)
% - "Population genetics, as currently understood, therefore, cannot chart the
%   evolution of an open system with indefinite variability." (Yedid, 2001)
% - "limited to abstract investigations of how information coding is influenced
%   by different selective pressures and mutational mechanisms" (Adami 2006)
% - "digital genomes are extremely short compared with even the smallest
%   free-living organisms and most viruses. Consequently, although the mutation
%   rate per genome is usually comparable to those of most RNA viruses92, the
%   rate per gene is unusually high." (Adami 2006)

% Advantage over models/simulations
% - "It could find a role, therefore, in verifying predictions made by
%   conventional theory in the context of an independent system of evolvable
%   self-replicators." (Yedid, 2001)
% - ME: From Yedid's 2001 manuscript... one of the biggest advantages of
%   an artificial life system is its open-endedness; models and simulations
%   of evolution are constrained by parameters and their ranges;
%   an evolving digital organism creates novelty and complexity
%   that cannot be forseen by people; it is not uncommon to build models
%   and simulations in order to understand general patterns in Avida
%   (pers. comm. Bjorn).
% - "Thus, even in this simplest of situations, the outcome of evolution can be
%   any of several genotypes, and it cannot be predicted which will emerge in
%   any particular realization." (Wilke 2002)
% - "Unlike other computational approaches to studying evolution (such as
%   genetic algorithms [1] or numerical simulations) digital organisms must
%   explicitly create a copy of their own GENOME (see Glossary) to reproduce,
%   and no particular genomic sequence is designated as the target or optimal
%   sequence. (Wilke 2002)
% - "By comparing results across wide ranges of parameter settings in the
%   digital world, as well as with results from biochemical organisms and from
%   mathematical theories, it is possible to disentangle general principles from
%   effects that are peculiar to a particular model organism." (Wilke, 2002)
% - "Darwinian mechanisms is inherently different from a simulation (such as a
%   genetic algorithm), because the fitness of a program is not determined a
%   priori by the user. Instead, as for biochemical life, those line- ages that
%   survive the competition for space, time and resources are the most fit — in
%   hindsight." (Adami 2006)

% Limitations of Avida
% - In a way, limitations depend on the question---you have to know which
%   questions to ask; otherwise, Avida won't help you
% - "research with digital organisms is restricted to abstract questions about
%   general principles." (Wilke 2002)
% - "design choices that enter the construction of the digital world
%   potentially influence the outcome of experiments, so care must be taken to
%   study only those questions for which this influence is expected to be
%   small." (Wilke 2002)
% - "Traditional theoretical or computational studies have the advantage that
%   the particular effect under investigation can often be described more
%   cleanly: Irrelevant details can be neglected in analytical calculations and
%   in custom-made computer simulations, whereas the dynamics of digital
%   organisms often become as messy as those of their biochemical counterparts.
%   (Wilke 2002)

% Other applications
% - Evolution of synchronization/desynchronization important in development
%   of communication protocols \citep{kno11}
% - cooperative information gathering (Beckmann 2007)
% - "Eventually, such robots, and the software that directs them, might evolve
%   without human interaction, at which point they would become part of the
%   ecosystem in which we live." (Wilke 2002)
% - "The alternative of evolving programs looked like a great idea to
%   Microsoft," [robust programs; also MS intersted in evolvable
%   programming languages] (O'Neil)
% - "says Lenski, but ‘biological principles, from evolution and genetics to
%   neurobiology and ecology, are informing computer scientists and engineers
%   in designing software and hardware ... and that holds tremendous promise
%   for the future’.(O'Neil)
% - "Evolutionary computation, a subfield of computer science, applies the
%   basic principles of genetic evolution to problem solving." (McKinley 2008)
% - digital evolution pro- vides an open-ended search technique for problems in
%   science and engineering. (McKinley 2008)
% - "Sys- tems must adapt to their environment, compensate for failures,
%   optimize performance, and protect themselves from attacks—all with minimal
%   human intervention" (McKinley 2008)
% - Applications to software design, new algorithms, cooperative communication
%   algorithms, minimizing energy consumption of mobile devices (McKinley 2008)
% - software resilience to varying conditions (Goldsby 2008)
% - robust distribution of data in a network (Knoester 2009)

%A detailed study of adaptation and speciation
%requires identifying individual alleles
%and tracking them through time, which is no easy task.
%%
%Computational and mathematical models of adaptation
%have been developed to study this process theoretically.
%%
%Models, however, miss some of the complexities of real, evolving organisms.



\section*{Study system: Avida}
\addcontentsline{toc}{section}{Study system: Avida}
\label{sec:avida}

In this section, we provide a brief overview of Avida;
for a full description, see \cite{ofr04}.
%
Avida is freely available at \url{http://avida.devosoft.org}.
%
In Avida, digital organisms are composed of
a linear sequence of instructions (akin to a haploid genome),
memory space in the form of registers and stacks,
pointers to memory locations,
and a central processing unit (CPU) that executes instructions.
%
The instruction set makes up an assembly-like programming language,
consisting of instructions for arithmetic operations,
memory manipulation (e.g., swapping registers or pushing into a stack),
conditional execution, iteration, input/output operations,
and allocation and copying of memory.
%
Organisms execute their instructions sequentially,
sometimes skipping instructions for conditional statements
or repeating the same instructions inside a loop;
when the last instruction is executed,
execution starts again at the first instruction.
%
By executing instructions in their genomes,
organisms are able to (1) replicate and (2) perform computational `tasks'
that increase the speed at which they replicate and thus increase fitness.



To replicate, an allocation instruction creates the memory space
required by the organism's offspring, and a copy instruction inside a loop
allows the organism to copy itself into the new memory space.
%
The copy instruction that allows organisms to replicate
has a configurable probability of making mistakes,
which introduces various kinds of mutations.
%
By default, replication is asexual.
%
However, Avida may be configured to perform sexual replication,
in which the genomes of two asexually-produced offspring are recombined
by exchanging two randomly-sized regions of their genomes.
%
The offspring (whether clonal or recombinants) are put into the population
in random locations, replacing whatever organisms were already there.
%
Generations are therefore overlapping, as offspring are born continuously,
replacing older individuals but who are likely not their parents.



In addition to replication, genomic instructions allow organisms
to acquire 32-bit input values and use them to perform computational tasks.
%
Tasks are boolean operations, such as NOT, AND, and OR,
and are applied to input values bit by bit.
%
For example, if input values were 8 bits,
the operation 10011101 AND 11101011 would produce 10001001
according to the rules of boolean logic for AND
(0~AND~0~=~0, 0~AND~1~=~0, 1~AND~0~=~0, and 1~AND~1~=~1.
%
In Avida, however, there is no AND operation nor any other boolean operation
except for NAND, from which all other boolean operations may be built,
a property of NAND known as `functional completeness' in boolean algebra.
%
For example, P~AND~Q, where P and Q are input values,
is equivalent to (P~NAND~Q)~NAND~(P~NAND~Q).
%
Therefore, in order to perform a task other than NAND,
digital organisms must make use of other instructions available,
which arise through mutation or recombination.



When an organism performs a task, the organism's `merit' is increased
by a specific amount, specified in a configuration file, for that task.
%
The merit of an organism is a unitless value used by Avida
to determine the number of instructions an organism may execute each time step.
%
If two organisms had the same merit,
they would execute the same number of instructions at each time step;
however, if one organism had twice the merit as another,
the first organism would execute twice the number of instructions
compared to the second in a single time step.
%
Thus, an organism with twice the merit as another
would replicate twice as fast.
%
Organisms initially inherit the merit of their parents;
otherwise, new organisms would be at a disadvantage
compared to the rest of the population.
%
The default environment rewards for nine binary (i.e., two-input) tasks.



Adaptation in Avida occurs naturally (i.e., it is not simulated),
as a result of the three ingredients required for natural selection:
inheritance, variation, and differential reproduction.
%
Inheritance comes from replication (sexual or asexual),
variation comes from mutation and recombination,
and differential reproduction comes from their rate of replication
(determined by their replication code and performance of tasks).
%
The ability to perform tasks evolves as organisms with the right mutations
replicate faster than others and therefore take over the population.
%
There are many ways in which to perform any one task,
and independently evolved organisms often evolve the same task
in different ways and with different degrees of efficiency.



\section*{Dissertation summary}
\addcontentsline{toc}{section}{Dissertation summary}

Evolutionary adaptation to a new environment
depends on the availability of beneficial alleles.
%
Beneficial alleles may appear as new mutations
or may come from standing genetic variation---%
alleles already present in the population
prior to the environmental change.
%
Adaptation from standing genetic variation
in sexually-reproducing populations
is expected to be faster than from new mutations
because beneficial alleles from standing genetic variation
occur at a higher starting frequency and are immediately available.
%
The distribution of fitness effects of alleles
from standing genetic variation are expected to be different
from that of new mutations because standing genetic variation
has been `pre-tested' by selection.
%
Whether adaptation uses standing genetic variation
or new mutations as a source of beneficial alleles is unknown.
%
In Chapter~\ref{chap:sgv} (p.~\pageref{chap:sgv}),
I conducted experimental evolution of digital organisms
to determine the source of beneficial alleles during adaptation.
%
I also tested the speed of adaptation
and the fitness effect of alleles
under these two sources of genetic variation.



Various processes may drive a biological population to acquire mutations
that reduce its fitness (i.e., ``deleterious mutations'').
%
A population with deleterious mutations may fully or partially recover
in fitness in two ways: reversion or compensatory adaptation.
%
Compensatory adaptation is the fixation of mutations
that ameliorate the effects of deleterious mutations
while the original deleterious mutations remain fixed.
%
Reversion is often the best way to recover and,
if a revertant mutation were to appear, the most probable route.
%
However, it has been found experimentally that
there are more compensatory mutations available
than the single revertant.
%
It has also been found that once compensatory adaptation
has begun, reversion becomes an increasingly improbable route
because the effect of the reversion changes with the genetic background.
%
Therefore, it seems that reversion has a limited
window of opportunity in which its full effect would be beneficial.
%
The fitness effect of the deleterious mutation,
the population size, and the mutation rate are three main factors
that will affect this window of opportunity.
%
In Chapter~\ref{chap:comp_rate} (p.~\pageref{chap:comp_rate}),
we used populations of digital organisms
to investigate the effect of these factors.
%
We found that the lower the initial fitness of the population,
the smaller the window of opportunity for reversion.
%
This result was partly caused by the stronger negative interactions
between compensatory mutations and the reversion.
%
We found that the window of opportunity for reversion
was highest the larger the population size,
but compensatory adaptation was most probable
at intermediate population sizes and lowest at the extremes.
%
Finally, we found that the higher the mutation rate,
the larger the window of opportunity reversion,
but it was smaller than expected because
the higher mutation rate caused more negative interactions
with compensatory mutations to occur.



Epistatic interactions among compensatory mutations that have
evolved in separate populations may form an intrinsic
postzygotic isolating barrier (i.e., hybrid inviability or sterility),
leading to biological speciation.
%
Indeed, about ten percent of genetic differences between species
comprise compensatory mutations.
%
In Chapter~\ref{chap:comp_spp} (p.~\pageref{chap:comp_spp}),
I tested experimentally whether compensatory adaptation
can lead to reproductive isolation (specifically, postzygotic isolation)
and whether it was more rapid and stronger
than in populations evolved through drift.
%
Surprisingly, the strength of this isolation was independent
of the effect size of the original deleterious mutations.
%
We also find that
both deleterious and compensatory mutations
contribute equally to reproductive isolation.
%
Our results suggest that compensatory adaptation may be
an important genetic mechanism of speciation,
and supports the view that intrinsic postzygotic isolation
can stem from intrinsic genetic mechanisms.



Reproductive isolation between populations often evolves as a byproduct of
independent adaptation to new environments, but the selective pressures of
these environments may be divergent (`ecological speciation') or uniform
(`mutation-order speciation').
%
In Chapter~\ref{chap:ecol_mo} (p.~\pageref{chap:ecol_mo}),
we directly compare the strength of reproductive isolation
(specifically, postzygotic) generated by
ecological and mutation-order processes.
%
We also tested the effect of gene flow as well as the dimensionality
(i.e., number of selective pressures) of the environments on the strength
of postzygotic isolation.
%
We found that ecological speciation generally formed stronger isolation than
mutation-order speciation, mutation-order speciation was more sensitive to gene
flow than ecological speciation, and environments with high dimensionality
formed stronger reproductive isolation than those with low dimensionality.
%
How various factors affect the strength of reproductive isolation
has been difficult to test in biological organisms, but the
use of artificial life, which provides its own genetic system that evolves,
allowed us to computationally test the effect of these factors more easily.



Under the Dob\-zhan\-sky-Mul\-ler model of speciation,
hybrid inviability or sterility results from the evolution
of genetic incompatibilities (DMIs) between species-specific alleles.
%
This model predicts that the number of pairwise DMIs between species
should increase quadratically through time,
but the few tests of this `snowball effect' have had conflicting results.
%
In Chapter~\ref{chap:snowball} (p.~\pageref{chap:snowball}),
we show that pairwise DMIs accumulated quadratically,
supporting the snowball effect.
%
The number of unfit hybrids has been proposed to accumulate faster
than quadratically because hybrids harbor more complex DMIs,
but we found that the accumulation was linear.
%
We show that more complex genetic interactions involved alleles
that rescued pairwise incompatibilities, explaining the discrepancy
between the accumulations of DMIs versus unfit hybrids.
%
Our results highlight the importance
of complex genetic interactions in speciation.

% Trick to hide bibliography from table of contents
\addtocontents{toc}{\protect\setcounter{tocdepth}{-1}}

\bibliographystyle{apalike}
\bibliography{intro}

\addtocontents{toc}{\protect\setcounter{tocdepth}{3}}
\addcontentsline{toc}{section}{Literature cited}

\end{doublespace}
